canaper: Categorical analysis of neo- and paleo-endemism in R

### 2022-09-30 ms/ms.Qmd
Update MS

---
title: "canaper: Categorical analysis of neo- and paleo-endemism in R"
bibliography: references.yaml
csl: methods-in-ecology-and-evolution.csl
highlight-style: github
crossref: 
  fig-title: Fig.
  tbl-title: Table
format:
  docx: default
  pdf:
    include-in-header:
      # keep figures from floating away
      text: |
        \usepackage{float}
        \makeatletter\renewcommand*{\fps@figure}{H}\makeatother
---

@@

Here, we present a new R package that implements CANAPE completely in R: `canaper`.
We strived to make `canaper` simple to use and efficient.
Parallel computing can be enabled with a single line of code.
`canaper` has undergone thorough code review through ROpenSci [@Boettiger2015] and is verified against a large number of unit tests (>95% coverage).
All results are reproducible by setting the random seed generator in R, in both sequential and parallel computing modes.

@@

The entire CANAPE workflow can be run with two functions, `cpr_rand_test()` and `cpr_classify_endem()`.
However, internally this entails several steps that the user should be aware of as follows.

@@

`canaper` does not include any plotting functions to visualize the results.
Rather, we recommend the `ggplot2` package or base R graphics to visualize results.
Here, we will demonstrate use of the `ggplot2` and `patchwork` packages to visualize the output of `canaper` (@fig-acacia-cpr).

@@

The color palette used in @fig-acacia-cpr, `mishler_endem_cols`, is the same as that used in @Mishler2014.
However, this palette may not be distinguishable to people with color-vision deficiency (CVD).
Alternate color palettes (e.g., `cpr_endem_cols`) are also vailable that include CVD-safe colors [@Okabe2002].
Palettes for plotting CANAPE and *p*-rank results can be selected with the `cpr_make_pal()` function.
The same plot visualized with CVD-safe colors is available in the online supporting material (Fig. S1).

@@

We are not aware of any other R packages that conduct the entire CANAPE pipeline automatically.
However, there is a large number of packages for analyzing species diversity [approximately 40 packages out of 15,300 as of 2019\; @Pavoine2020].
Some of those that are more closely related to `canaper` include the following.
`picante` [@Kembel2010] was one of the first packages to offer calculation of phylogenetic alpha diversity such as MPD, MNTD, and PD [@Webb2000].
`phyloregion` [@Daru2020a] implements sparse matrix encoding to increase computing efficiency of PD, and is used by `canaper`.
`vegan` [@Oksanen2017] performs a wide range of non-phylogenetic community ecology analyses.
`vegan` includes by far the greatest number of algorithms for generating random communities (> 30) and is used by `canaper`.
`adiv` [@Pavoine2020] provides a flexible framework for analyzing alpha and beta diversity of biological communities based on either phylogenetic or trait distances.
`PhyloMeasures` [@Tsirogiannis2016] implements efficient routines for calculating phylogenetic diversity metrics.
`EcoSimR` [@Gotelli2013a] provides randomization algorithms as well as functions for characterizing community matrices and null distributions.
`PhyloMeasures` and `EcoSimR` have both been removed from CRAN and may not be currently maintained.

@@

All data and code are available in the `canaper` R package, which is distributed by the Comprehensive R Archive Network (CRAN).
Source code is available on GitHub (<https://github.com/ropensci/canaper>) and Zenodo (<https://doi.org/10.5281/zenodo.5094032>).
Usage of `canaper` is documented at <https://docs.ropensci.org/canaper/>.
The `acacia` dataset was originally published by @Mishler2014.

@@

@@

**@fig-acacia-cpr.** Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*.
(a) Map of Australia showing grid-cells (communities) colored by endemism type.
(b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs. a comparison tree with all non-zero branch lengths set to equal length, colored according to endemism type. This figure reproduces Figure 3 of @Mishler2014.

@@

>>>>

---
title: "canaper: Categorical analysis of neo- and paleo-endemism in R"
bibliography: references.yaml
csl: methods-in-ecology-and-evolution.csl
highlight-style: github
crossref: 
  fig-title: Fig.
  tbl-title: Table
format:
  docx:
    reference-doc: custom-reference.docx
  pdf:
    include-in-header:
      # keep figures from floating away
      text: |
        \usepackage{float}
        \makeatletter\renewcommand*{\fps@figure}{H}\makeatother
---

@@

Here, we present a new R package that implements CANAPE completely in R: `canaper`.
We strived to make `canaper` simple to use and efficient.
Parallel computing can be enabled with a single line of code.
`canaper` has passed code review meeting the "silver" standards for statistical software at ROpenSci [@Boettiger2015], and is verified against a large number of unit tests (>99% coverage).
All results are reproducible by setting the random seed generator in R, in both sequential and parallel computing modes.

@@

The entire CANAPE workflow can be run with two functions, `cpr_rand_test()` and `cpr_classify_endem()`.
However, internally this entails several steps that the user should be aware of as follows (@fig-workflow).

@@

`canaper` does not include any plotting functions to visualize the results.
Rather, we recommend the `ggplot2` package or base R graphics to visualize results.
Here, we demonstrate use of the `ggplot2` and `patchwork` packages to visualize the output of `canaper` (@fig-acacia-cpr).

@@

The color palette used in @fig-acacia-cpr, `mishler_endem_cols`, is the same as that used in @Mishler2014.
However, this palette may not be distinguishable to people with color-vision deficiency (CVD).
Alternate color palettes (e.g., `cpr_endem_cols`) are also available in `canaper` that include CVD-safe colors [@Okabe2002].
Palettes for plotting CANAPE and *p*-rank results can be selected with the `cpr_make_pal()` function.
The same plot visualized with CVD-safe colors is available in the online supporting material (Fig. S1).

@@

We are not aware of any other R packages that conduct the entire CANAPE pipeline automatically.
However, there is a large number of packages for analyzing species diversity [approximately 40 packages out of 15,300 as of 2019\; @Pavoine2020].
Some of those that are more closely related to `canaper` include the following.
`picante` [@Kembel2010] was one of the first packages to offer calculation of phylogenetic alpha diversity such as MPD, MNTD, and PD [@Webb2000].
`phyloregion` [@Daru2020a] implements sparse matrix encoding to increase computing efficiency of PD, and is used by `canaper`.
`vegan` [@Oksanen2017] performs a wide range of non-phylogenetic community ecology analyses.
`vegan` includes by far the greatest number of algorithms for generating random communities (> 30) and is used by `canaper`.
`adiv` [@Pavoine2020] provides a flexible framework for analyzing alpha and beta diversity of biological communities based on either phylogenetic or trait distances.
`PhyloMeasures` [@Tsirogiannis2016] implements efficient routines for calculating phylogenetic diversity metrics.
`EcoSimR` [@Gotelli2013a] provides randomization algorithms as well as functions for characterizing community matrices and null distributions.
As of writing, `PhyloMeasures` and `EcoSimR` have both been removed from CRAN and may not be currently maintained.

@@

All data and code are available in the `canaper` R package, which is distributed by the Comprehensive R Archive Network (CRAN).
`canaper` source code is available on GitHub (<https://github.com/ropensci/canaper>) and Zenodo (<https://doi.org/10.5281/zenodo.5094032>).
Usage of `canaper` is documented at <https://docs.ropensci.org/canaper/>.
Code and data used to run analyses and generate this manuscript available at <https://github.com/joelnitta/canaper_ms>.
The `acacia` dataset was originally published by @Mishler2014.

@@

**@fig-workflow.** `canaper` workflow and functions to conduct categorical analysis of neo- and paleo-endemism (CANAPE).
Input data are a community matrix and phylogeny.
First, the user selects a randomization algorithm and its settings, with optional assistance from the `cpr_iter_sim()` function.
Next, a set of random communities are generated, various metrics (e.g., phylogenetic endemism, relative phylogenetic endemism) are calculated for the input community and each random community, and the rank (*p*-value) of each observed metric is compared to those of the random communities by the `cpr_rand_test()` function.
Finally, endemism types are categorized on the basis of the ranked metrics with the `cpr_classify_endem()` function.
For details on metrics and classification, see "Analysis Workflow".
For details on functions, see "Major Functions".

@@

**@fig-acacia-cpr.** Categorical analysis of neo- and paleo-endemism (CANAPE) of Australian *Acacia*.
(a) Map of Australia showing grid-cells (communities) colored by endemism type.
Latitude and longitude projected into equal area Australian Albers (EPSG:3577) coordinate system [@Butler2007]; units in m.
(b) Scatterplot of comparing phylogenetic endemism (PE) of each community as measured on the original tree vs. a comparison tree with all non-zero branch lengths set to equal length, colored according to endemism type. This figure reproduces Figure 3 of @Mishler2014.

@@

![`canaper` workflow and functions to conduct categorical analysis of neo- and paleo-endemism (CANAPE)](workflow.png){#fig-workflow}
